The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-05-30 paper/ms.tex
results section

A quantitative description of the \ly line has been reached through
Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. In the last
two decades these studies have become popular due to the
availability of computing power. Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}. Later on, the effects of clumpy media
\citep{Hansen06} and expanding/contracting shell/spherical geometries started to
be studied \citep{Verhamme06,Dijkstra06,Ahn2014}. Similar codes have applied
these results to semi-analytic models of galaxy formation \citep{Orsi12, Garel2012} and
results of large hydrodynamic simulations
\citep{CLARA,Forero12,Behrens13}. Recently, Monte Carlo codes have also
been applied to the results of high resolution hydrodynamic
simulations of individual
galaxies \citep{Laursen09,Barnes11,Verhamme12,Yajima12}. Meanwhile, recent
developments have been focused on the systematic study of clumpy
outflows \citep{DijkstraKramer}and anisotropic velocity configurations
\citep{Zheng2013}.

@@

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results in the
interpretation of LAEs observations with high resolution
spectroscopy. In the last section we present our conclusions.

@@

We use the full width at half maximum (FWHM) to quantify the line
broadening. We measure this width from the velocity histogram by
finding the values of the velocities at half maximum. We use lineal
interpolation between histogram points to get a more precise value for
this width.

Figure \ref{fig:widthvsmu} shows the FWHM for all models as a function
of the viewing angle. The FWHM increases for decreasing values of
$\mu$ increases and increasing values of $V_{\rm max}$. In Figure
\ref{fig:widthvsvelocity} we fix $|\mu|<0.1$ and show the FWHM as a
function of rotational velocity. For this case we parametrize the
dependency of the line width with  $V_{\rm max}$ as

\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the velocity width in the static case and $\lambda$ 
is a positive scalar to be determined as a fit to the data.

With this test we want to know to what extent the new velocity width
can be expressed as a quadratic sum of the two relevant velocities in
the problem. We fit simultaneously all the points corresponding to
central and homogeneous models to find $\lambda_{\rm c} = 0.84 \pm
0.06$ and $\lambda_{\rm h}= 0.54\pm 0.1$ respectively.

@@

We measure the peak maxima position to quantify the transition from
double into single peak. We do this as a function of the viewing angle
$\theta$ and the maximum rotational velocity. The results are
summarized in Figure  \ref{fig:maximumsvsvelocity}. There are two
interesting features that deserve attention. First, for a viewing angle
parallel to the rotational axis ($\mu\sim 1.0$) the maxima of all
models are similar regardless of the rotational velocity. Second, at a
viewing angle perpendicular to the rotation axis ($\mu\sim 0.0$) there
is a good fraction of models that become single peaked, a feature that
appears more frequently for homogeneously distributed sources when all
the other parameters are equal.

@@

We now estimate the escape fraction $f_{\rm esc}$ for the dusty models. Our main
result is that we do not find any dependence with the viewing angle
but there is a strong dependence with the rotational velocity for the
homogeneous models.

The left panel in Figure \ref{fig:escape_fraction} summarizes our
results. The left panel shows the escape fraction as a function of the
rotational velocity. To construct this Figure we take into account all
the photons regardless of its outgoing direction. We observe that the
curves for the central source distribution stay flat, while for the
homogeneous case there is a clear rise with rotational velocity.
Rotation has a higher relative impact in the models with low optical
depth. For instance, in models with $\tau_{\rm H}=10^5$, the static
escape fraction is $0.26$ and increases to $0.37$ for $V_{\rm
  max}=300$ \kms. Table \ref{table:escape} lists all the values for
the escape fraction.

@@

A possible explanation for these two results is that some photons
close to the surface can escape with a low number of scatterings. This
has a two-fold consequence: the escaping photons do not have time to
scatter away from the line's center and also have less chances to find
an absorbing dust grain.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the cosinus of the
outgoing angle $|\cos\theta|$ and the rotational velocity
$V_{\rm max}$. From the right panel observe that the number of
scatterings and the outgoing angle are independent. This plot
corresponds to the specific case of the central model with $\tau=10^5$ and
$V_{\rm max}=300$\kms; but we have veryfied that this holds for all
models. This allows us to focus to use all the photons in the
simulations regardless of their value of $\cos\theta$.

In the right panel of Figure \ref{fig:Nscatt} we observe how the
average number of scatterings decreases for larger rotational
velocities in the case of homogeneous source distributions. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case. For
the central source distribution the average number of scatterings
$\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$.

In static cases, a large value of the optical depth correlates with a
high number of scatterings.  This can be precisely quantified in the
static slab. In this model for centrally emitted sources the average
number of scatterings depends only on the optical depth $\langle
N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}, in the case of homogeneously distributed
sources $\langle N_{\rm   scatt}\rangle=1.16\tau_{\rm   H}$
\citep{Harrington73}.

We perform a similar quantification of in our speherical setup for the
scatic spheres. We find that for the central model the number of
scatterings is proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively.
For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this represents
factors of $(0.66, 0.59, 0.55)$ lower than the centrally  emitted photons.

@@

In order to gain a deeper understanding of these results  we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we these 
results in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$.

The top right panel of Figure \ref{fig:Nscatt2D} (homogeneous sources,
high rotational velocity) supports our hypothesis about
the photo-sphere in the homogeneous distribution. In this case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This also explains the decrease in the average number of
scatterings observed in Figure \ref{fig:Nscatt}.

However, for a central distribution the situation is quite different
(lower panels). In this case the number of scatterings remains high,
in the order of the optical depth, but the two peaks do get closer to each
other. Here the most probable physical picture is that each
scattering, due to the bulk velocity of the gas, is inefficient in
driving the photon outside the line center.

>>>>

A more detailed quantitative description of the \ly line has been reached through
Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. In the last
two decades these studies have become popular due to the
availability of computing power. Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}. Later on, the effects of clumpy media
\citep{Hansen06} and expanding/contracting shell/spherical geometries started to
be studied \citep{Verhamme06,Dijkstra06,Ahn2014}. Similar codes have applied
these results to semi-analytic models of galaxy formation \citep{Orsi12, Garel2012} and
results of large hydrodynamic simulations
\citep{CLARA,Forero12,Behrens13}. Recently, Monte Carlo codes have also
been applied to the results of high resolution hydrodynamic
simulations of individual
galaxies \citep{Laursen09,Barnes11,Verhamme12,Yajima12}. Meanwhile, recent
developments have been focused on the systematic study of clumpy
outflows \citep{DijkstraKramer}and anisotropic velocity configurations
\citep{Zheng2013}.

@@

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results. In
the last section we present our conclusions.

@@

We use the full width at half maximum (FWHM) to quantify the line
broadening. We measure this width from the velocity histogram by
finding the values of the velocities at half maximum intensity. We use
lineal interpolation between histogram points to get a value more precise
than the bin size used to construct the line intensity.

Figure \ref{fig:widthvsmu} shows the FWHM for all models as a function
of the viewing angle. The FWHM increases for decreasing values of
$\mu$ and increasing values of $V_{\rm max}$. In Figure
\ref{fig:widthvsvelocity} we fix $|\mu|<0.1$, i.e. viewing angle
perpendicular to the rotation axis, and plot the FWHM as a
function of rotational velocity.

we parametrize the dependency of the line width with  $V_{\rm max}$ as
%
\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the velocity width in the static case and $\lambda$ 
is a positive scalar to be determined as a fit to the data.  With this
test we want to know to what extent the new velocity width can be
expressed as a quadratic sum of the two relevant velocities in the
problem. That all the models fall into two families of straight lines,
for central and homogeneously emitted sources, justifies this choice
of parameterization.

We fit simultaneously all the points corresponding to each one of this
families to find $\lambda_{\rm c} = 0.84 \pm 0.06$ and $\lambda_{\rm
  h}= 0.54\pm 0.10$ respectively.

@@

We measure the peak maxima position to quantify the transition from
double into single peak profiles. We do this as a function of the viewing angle
$\theta$ for different rotational velocities. The results are
summarized in Figure  \ref{fig:maximumsvsvelocity}. There are two
interesting features that deserve attention. First, for a viewing angle
parallel to the rotational axis ($\mu\sim 1.0$) the maxima of all
models are similar regardless of the rotational velocity. Second, at a
viewing angle perpendicular to the rotation axis ($\mu\sim 0.0$) a
good fraction of models become single peaked. This feature appears
more frequently for homogeneously distributed sources if all the
other parameters are equal.

@@

We now estimate the escape fraction $f_{\rm esc}$ for the dusty
models. The main  result is that we do not find any dependence
with the viewing angle but there is a strong dependence with the
rotational velocity for the homogeneous models.

The left panel in Figure \ref{fig:escape_fraction} summarizes the
findings about the dependence with the rotational velocity. The left
panel shows the escape fraction as a function of the rotational
velocity. To construct this Figure we take into account all the
photons regardless of its outgoing direction. We observe that the 
curves for the central source distribution stay flat, while for the
homogeneous case there is a clear rise with rotational velocity.
Rotation has a higher relative impact in the models with low optical
depth. For instance, in models with $\tau_{\rm H}=10^5$, the static
escape fraction is $0.26$ and increases to $0.37$ for $V_{\rm
  max}=300$ \kms. Table \ref{table:escape} lists all the values for
the escape fraction.

@@

A possible explanation for these two results is that some photons
close to the surface can escape with a low number of scatterings. This
has a two consequences: the escaping photons do not have time to
scatter away from the line's center and also have less chances to find
an absorbing dust grain.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the cosinus of the
outgoing angle $|\cos\theta|$ and the rotational velocity
$V_{\rm max}$. From the right panel observe that the number of
scatterings and the outgoing angle are independent. This plot
corresponds to the specific case of the central model with $\tau=10^5$ and
$V_{\rm max}=300$\kms, but we have veryfied that this holds for all
models. This allows us to construct statistics using all the outgoing
photons regardless of their value of $\cos\theta$.

The right panel of Figure \ref{fig:Nscatt} shows how the
average number of scatterings decreases for larger rotational
velocities in the case of homogeneous source distributions. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case. For
the central source distribution the average number of scatterings
$\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$.

In order to gain a deeper understanding of these results  we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show these 
results in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$.

The top right panel of Figure \ref{fig:Nscatt2D} (homogeneous sources,
high rotational velocity) supports our hypothesis about
the photo-sphere in the homogeneous distribution. In this case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This also explains the decrease in the average number of
scatterings observed in Figure \ref{fig:Nscatt}. We have also veryfied
that these photons are indeed emitted closer to the surface than to
the center of the sphere.

However, for a central distribution the situation is quite different
(lower panels in Figure \ref{fig:Nscatt2D}). In this case the number
of scatterings remains high, in the order of the optical depth, but
the two peaks do get closer to each other. The most probable
physical picture is that each scattering, due to the bulk velocity of
the gas, is inefficient in driving the photon outside the line
center.

We now proceed with a more detailed quantification of the number of
scatterings in our simulations. We guide the measurement by the fact
that in static cases, a large value of the optical depth correlates
with a high number of scatterings.  This has been precisely quantified
in the case of static infinite slab. In this model for centrally
emitted sources the average number of scatterings depends only on the
optical depth $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}, for homogeneously distributed sources
$\langle N_{\rm   scatt}\rangle=1.16\tau_{\rm   H}$
\citep{Harrington73}.

We perform a similar quantification of in our speherical setup for the
static spheres. We find that for the central model the number of
scatterings is proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively.
For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this represents
factors of $(0.66, 0.59, 0.55)$ lower than the centrally  emitted photons.

@@
